---
title: "DATA 607 Project 1"
subtitle: "Chess ELO"
author: 'Donny Lofland'
date: '9/14/2019'
output:
  html_document:
    theme: cerulean
    highlight: pygments
    css: ./lab.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stringr)
```

Source files: [https://github.com/djlofland/DATA607_F2019/tree/master/Project1](https://github.com/djlofland/DATA607_F2019/tree/master/Project1)

## Project

In this project, you’re given a text file with chess tournament results where the information has some structure. Your
job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database)
with the following information for all of the players:

Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents

For the first player, the information would be:

> Gary Hua, ON, 6.0, 1794, 1605

1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and
dividing by the total number of games played.

If you have questions about the meaning of the data or the results, please post them on the discussion forum. Data
science, like chess, is a game of back and forth…

The chess rating system (invented by a Minnesota statistician named Arpad Elo) has been used in many other contexts,
including assessing relative strength of employment candidates by human resource departments.

You may substitute another text file (or set of text files, or data scraped from web pages) of similar or greater
complexity, and create your own assignment and solution. You may work in a small team. All of your code should be in
an R markdown file (and published to rpubs.com); with your data accessible for the person running the script.

## Load the Dataset

```{r load_data}

# Load the chess ELO document into R, skip over the 1st 4 header row, we won't need those
elo_data <- read_lines('tournamentinfo.txt', skip=4)

# quick sanity check to see if the data loaded
head(elo_data, n=9)

# Let see how many rows so we can verify later munging steps - should be 3 rows per player
print(paste('Rows of Raw Data:', length(elo_data)))
print(paste('Players found:', length(elo_data) / 3))
```

## Munge and Extract Data

```{r}
# Remove every 3rd line with dashes
elo_data <- elo_data[-seq(3, length(elo_data), by=3)]

# for each player, combine their 2 rows of data into 1 wide row
elo_data <- paste(elo_data[seq(1, length(elo_data), by=2)], elo_data[seq(2, length(elo_data), by=2)], sep="")

# How many player rows are still here - should match count above
print(paste('Players found:', length(elo_data)))
```

### Convert into a DataFrame

```{r}
elo_list <- str_split(elo_data, "\\|")

elo_df <- NULL

for (r in 1:length(elo_list)) {
  trim_row <- unlist(lapply(elo_list[[r]], str_trim))
  elo_df <- rbind(elo_df, trim_row)
}

elo_df <- data.frame(elo_df)
elo_df <- as_tibble(elo_df)

elo_df
```

### Fix DF Column names

Lets calculate the number of opponent matches were played.  While our dataset had a fixed 7 opponents. maybe in the future, a contest fewer or more match columns provided.

```{r}
# Figure out how many opponent matches occurred
total_cols <- ncol(elo_df)                      # parsed columns found
fixed_cols <- 7                                 # "id", "name", "points", "state", "score", "num", "tail" 
opponent_cols <- (total_cols - fixed_cols) / 2  # there are 2 columns per opponent

# Setup list of column names we'll appy to our DF
opp_ids <- list()     # we'll also use this list later when cleaning up columns
opp_colors <- list()  # we'll be deleting these columns late

# Build lists of opponent related columns
for (i in 1:opponent_cols) {
  opp_ids <- c(opp_ids, paste("oppID_", i, sep=""))
  opp_colors <- c(opp_colors, paste("oppColor_", i, sep=""))
}

# Now set column names on our DF
colnames(elo_df) <- c("id", "name", "points", opp_ids, "state", "score", "num", opp_colors, "trail")

names(elo_df)
```

### Drop Unnecessary Columns

```{r}
# drop unnecessary columns
keep <- c("id", "name", "points", "state", "score", unlist(opp_ids))
elo_df <- elo_df[keep] 

names(elo_df)
```

### Clean up data in each column

```{r}
# Lets clean up our columns (fix datatypes and extract parts where necessary)
elo_df$id <- as.integer(lapply(elo_df$id, as.character))
elo_df$name <- unlist(lapply(elo_df$name, as.character))
elo_df$points <- as.numeric(elo_df$points)

# The players pre-competition score is buried in the $score string, use regex to pull out
elo_df$score <- unlist(lapply(elo_df$score, as.character))
elo_df$score <- as.integer(str_match(as.character(elo_df$score), ".*R: *(\\d+).*")[,2])

# Loop thru the opponent columns and extract the opponent ID's.  We'll need those to look up and average opponent scores later
for (i in 1: length(opp_ids)) {
  elo_df[[opp_ids[[i]]]] <- unlist(lapply(elo_df[[opp_ids[[i]]]], as.character))
  elo_df[[opp_ids[[i]]]] <- as.integer(str_match(elo_df[[opp_ids[[i]]]], ".+ (\\d+)$")[,2])
}

str(elo_df)
```

## Calculate player's Average Opponent Score

```{r}
# I kow there is a way to do this vectorized, probably with nested lappy(), but after many hours of head banging, I decided loops were good enough for this dataset.
rate_cols <- list()

# Loop thru each oppID_* column
for (i in 1:length(opp_ids)) {
  # Create the column name that 'll hold the opponent rating
  rate_col <- paste("oppRating_", i, sep="")
  rate_cols <- c(rate_cols, rate_col)
  
  # Create a new rating column (seed with id, we'll replace this with rating in the next step)
  elo_df[rate_col] <- elo_df[opp_ids[[i]]]
  
  # Loop thru all cells in the new rating column, look up the score for the id, then overwrite
  # note, I'm make sure the ID isn't larger than our row count as a safety value
  for (r in 1:nrow(elo_df)) {
    if(!is.na(elo_df[[r, rate_col]]) && elo_df[[r, rate_col]] <= nrow(elo_df)) {
       elo_df[[r, rate_col]] <- as.integer(elo_df[elo_df$id==elo_df[[r, rate_col]], 'score'])
    }
  }
}

# with Opponent scores populated, the mean is easy.  Note that I left NA's in place on purpose so I could do quick mean calcualtion with na.rm=TRUE.
elo_df$avg_opp_ratings <- rowMeans(elo_df[unlist(rate_cols)], na.rm=TRUE)
elo_df$avg_opp_ratings
```

```{r}
rowMeans(elo_df[unlist(rate_cols)], na.rm=TRUE)

```


## Final Output

### Prepare DF for export

```{r}
final <- elo_df[c('name', 'state', 'points', 'score', 'avg_opp_ratings')]
final
```

### Save CSV file

```{r}
write.csv(final, file="elo_export.csv")
```



## What I Learned

- Python is soooo much easier to work.  Fortunately, I'm already bald so didn't tear any hair out. 
-
- 
